---
title: "FinalProject"
author: "Stephen Diehl"
date: "June 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This is the final project for the Data Science Specialization Course: Practical Machine Learning.

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

## Reading in the Data

Looking at the data, there are 3 ways a number can be NA.  Account for that in read_csv.
```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(caret)
pml_train <- read_csv("./data/pml-training.csv", na=c("NA", "", "#DIV/0!"))
pml_test <- read_csv("./data/pml-testing.csv", na=c("NA", "", "#DIV/0!"))
print(paste("Train", paste(dim(pml_train), collapse=" ")))
print(paste("Test", paste(dim(pml_test), collapse=" ")))
```

## Cleaning the Data

```{r}
# get the counts of NA in each column
na_count <-sapply(pml_train, function(y) sum(length(which(is.na(y)))))

# if there are more than 95% NAs, remove that column
cols_to_remove <- which(na_count > 0.95*19622)

# examination of columns by eye shows that first 6 columns are also not meaninful predictors
cols_to_remove <- union(cols_to_remove, 1:6)

# whatever is done to one data set must be done to the other
train <- subset(pml_train, select = -cols_to_remove)
test <- subset(pml_test, select = -cols_to_remove)
print(paste("Train", paste(dim(train), collapse=" ")))
print(paste("Test", paste(dim(test), collapse=" ")))
```

Normally data would be split something like 60% train, 20% validation, 20% test.  Here we see Train has 19622 records and Test only has 20 records.  So pml_test is not really a test set.  It will be used to submit results for the last quiz in the course, but that's it.

Break pml_train into train and test.
```{r}
set.seed(314159) # For reproducibile purpose
indexes <- createDataPartition(pml_train$classe, p=0.70, list=F)
train <- pml_train[indexes, ]
test <- pml_train[-indexes, ]
```

## Building the Model

Random Forests work well with almost no preprocessing.  Its okay to keep columns that are mostly NA, or rows that have NA in them.  This does not affect the accuarcy of the Random Forest.

That said, if a column cannot possibly be of value as a predictor, such as:

- Record number in column 1
- Time value in column 3
- Time value in column 4
- Time value in column 5

then these columns should be removed.  Whatever is done to the training data set, must also be done to the other data sets.

Also, let's remove columns with near zero variance.

```{r}
indexes <- union(nearZeroVar(pml_train), c(1,3,4,5))

train_processed <- subset(train, select = -indexes)
test_processed <- subset(test, select = -indexes)
test_quiz_processed <- subset(pml_test, select = -indexes)
print(paste("Train", paste(dim(train_processed), collapse=" ")))
print(paste("Test",  paste(dim(test_processed), collapse=" ")))
print(paste("Quiz",  paste(dim(test_quiz_processed), collapse=" ")))
```

```{r}
controlRF <- trainControl(method="cv", 5)
modelRF <- train(classe ~ ., data=train_processed, method="rf", trControl=controlRF, 
                 ntree=250, na.action = na.pass) # NA's are okay
modelRF
```

